---
title: "Predictive performance of presence-only species distribution models: a benchmark study with reproducible code. Ecological Monographs."
description: |
  Appendix S2
author:
  - name: Roozbeh Valavi 
    url: https://github.com/rvalavi
    affiliation: The University of Melbourne, Australia
    affiliation_url: https://ecosystemforest.unimelb.edu.au/
    orcid_id: 0000-0003-2495-5277
  - name: Gurutzeta Guillera-Arroita 
    affiliation: The University of Melbourne, Australia
    affiliation_url: https://ecosystemforest.unimelb.edu.au/
    orcid_id: 0000-0002-8387-5739
  - name: José J. Lahoz-Monfort 
    affiliation: The University of Melbourne, Australia
    affiliation_url: https://ecosystemforest.unimelb.edu.au/
    orcid_id: 0000-0002-0845-7035
  - name: Jane Elith 
    affiliation: The University of Melbourne, Australia
    affiliation_url: https://ecosystemforest.unimelb.edu.au/
    orcid_id: 0000-0002-8706-0326
twitter: 
  site: "@ValaviRoozbeh"
  creator: "@ValaviRoozbeh"
journal: 
  title: "Ecological Monographs"
  pasge: "1-27"
  # issn: 2490-1752
  # publisher: ESA
# volume: 44
# issue: 4
doi: "10.1002/ecm.1486"
date: "2021-10-6"
bibliography: references.bib
bib-humanities: true
output: 
  distill::distill_article:
    toc: true
    toc_depth: '2'
    theme: theme.css
creative_commons: CC BY
header-includes: 
  \usepackage{caption}
  \renewcommand{\figurename}{Fig.} 
  \renewcommand{\thefigure}{S\arabic{figure}} 
  \renewcommand{\thetable}{S\arabic{table}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

    
# Introduction

In this document, we show how different **species distribution models (SDMs)** can be fitted to **presence-background** data in the R programming language. Several statistical and machine learning algorithms are presented here, covering all methods presented in the main text. Each model is fitted using the same code we used for producing the results in our paper. Some additional suggestions are also provided that could be useful for smaller modeling exercises, but that we could not apply due to the large number of species and computation limits. The code provided here can be used for modeling any binary response.

```{r}
library(raster)
```


